Distributional Trees and Forests for Circular Data with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-04 pkg/circtree/inst/wind_paper/draft/draft.Rnw
first draft of result description
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2811 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b9943e69d7ce0f93299aabf666976eb308c1e76f

\subsection{Motivational Example}
As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

@@

\subsection{Results}
Ibk/Vie, 1h/3h

\section{Discussion and Outlook}

Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.

>>>>

\subsection{Motivational Example}\label{sec:mot_ex}
As a first illustration Figure~\ref{fig:tree_ibk} displays a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

@@

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour(s).
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction rapidly and almost randomly they are not 
expected to have a meaningful impact on predictions. Moreover, the performance and the stability 
especially of the employed reference models might suffer from the influence of these ``noise'' 
observations. Therefore, only observations corresponding to a wind speed of at least $1$ m/s are 
considered.

@@

\subsection{Results}
In order to test and compare the models over varying settings not only 1-hour but also 3-hour
forecasts are considered at both airports yielding four different scenarios of which the results
are discussed in this section.

Figure~\ref{fig:boxplot_crpsraw} illustrates the CRPS values of the employed models with a lower 
CRPS indicating better performance. 
In all four settings it can be observed that on average the distributional forest provides the best 
fitting probabilistic predictions followed by the distributional tree. Especially in Vienna the 
persistence (hour) model is only slightly behind while the linear model and the climatology
show clearly higher CRPS values.

Comparing the performances over the different locations the narrower boxes in Vienna represent
a lower variation. This can be explained by more abrupt changes of wind directions in Innsbruck
due to the surrounding mountains channeling wind and in that way allowing only for a limited
number of possible wind directions (mainly the three wind regimes as discussed in the motivational
example in Section~\ref{sec:mot_ex}). For that reason, relatively stable models such as persistency
produce well fitting forecasts as long as wind directions remain in one of these possible states,
however, once a change takes place it can only be an abrupt and major shift leading to high
prediction errors as supported by the wide boxes of the persistence model in Innsbruck. In Vienna,
on the other hand, changes can also appear in a smaller extend as the geographic surroundings do not
constrain wind directions. This wider range of possible observations might also weaken the performance
of the climatology as expressed by the higher CRPS values in Vienna.

The impact of the different forecasting time windows of 1 and 3 hours can hardly be noticed, especially
as it does not seem to effect the order of which method performs best. But taking a closer look
at the CRPS values of the tree, forest and persistency reveals lower scores and hence a better 
performance for the 1-hour forecasts as it could have been expected due to increasing uncertainties
in a longer time window.

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
\includegraphics{boxplot_crpsraw.pdf}
\caption{}
\label{fig:boxplot_crpsraw} 
\end{figure}

With the aim of excluding any possible influence of the selection of testing and training data
the performances of the methods are also compared based on the CRPS skills score as illustrated in 
Figure~\ref{fig:boxplot_crpsskill}. Using the climatology as reference model the values on the y-axis
indicate the percentage of improvement of each model compared to the reference model. Overall
the same conclusions can be drawn as from Figure~\ref{fig:boxplot_crpsraw} with the distribution
forest leading to the highest overall improvement in model fit.

\begin{figure}[t]
\centering
\setkeys{Gin}{width=1.0\textwidth}
%\hspace{-2cm}
\includegraphics{boxplot_crpsskill.pdf}
\caption{}
\label{fig:boxplot_crpsskill} 
\end{figure}

\section{Discussion and Outlook}
Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is particularly
useful for modeling wind direction in mountainous terrain where wind shifts
can occur due to turns of the pressure gradients along a valley.
